Experimentally informed site-specific substitution models deepen phylogenetic estimates of the divergence of viral lineages

### 2018-02-14 manuscript/manuscript.tex
small tweaks to model feature section

\begin{abstract}
\noindent  
Molecular phylogenetics is often used to estimate the time since the divergence of modern gene sequences.
Such phylogenetic techniques often estimate substantially shallower divergence times than other methods.
For instance, in the case of viruses there is independent evidence that molecular phylogenetics can underestimate deep divergence times.  
This discrepancy is thought to be caused in part by inadequate models of purifying selection leading to branch-length underestimation.
Here we show that substitution models informed by experimental measurements of the purifying selection due to site-specific amino-acid preferences lengthen the deep branches on phylogenies of influenza virus hemagglutinin.
This deepening of branch lengths is due to better modeling of the stationary state of the substitution models, and is independent of the branch-length-extension that results from modeling site-to-site variation in substitution rate.
The deepening of branch lengths from experimentally informed site-specific substitution models is similar to that achieved by other approaches that allow the stationary state to vary across sites.
However, the improvements from these site-specific models are limited by the inherent tension between the enhanced accuracy of accounting for site-specific amino-acid preferences and the fact that these preferences shift over long evolutionary times.
Overall, our work underscores the importance of modeling how site-specific purifying selection affects the stationary state when estimating deep divergence times. 
\end{abstract}

@@

Proteins evolve under both purifying selection to maintain their structure and function. 
This purifying selection is not homogenous across sites in a protein.
It is also not homogenous across the different amino acids at a given site.
For instance, some protein sites strongly prefer hydrophobic amino acids, others may be constrained to just one or a few amino acids, and yet others may tolerate many amino acids.
In general, these constraints are highly idiosyncratic among sites, and so pose a challenge for phylogenetic substitution models.

The most common strategy to model rate variation is to allow the rate of non-synonymous change to vary among sites following some statistical distribution.
For example, in the M5 of \cite{yang2000codon} the implied dN/dS parameter, $\omega$, follows a gamma distribution (GY94+$\Gamma\omega$, \ref{fig:model_feature}A).
Under the GY94+$\Gamma\omega$ and other such models, some sites have a higher rate of non-synonymous change than others. 
However, this rate is agnostic to the amino-acid identities themselves, and so all non-synonymous changes at at given site are treated equally.

In contrast, so-called ``mutation-selection'' models~\citep{halpern1998evolutionary} account for purifying selection by explicitly defining a different set of amino-acid preferences at each site in the protein. 
This more mechanistic formulation, without an arbitrary statistical distribution, results in a site-specific stationary state (\ref{fig:model_feature}B). 
These models better capture the site-to-site differences in amino-acid composition that is an obvious features of real proteins, and indeed they generally better describe actual evolution than models with only rate variation~\citep{lartillot2004bayesian, le2008phylogenetic, rodrigue2010mutation,hilton2017phydms,bloom2014experimentally}.
The specificity of mutation-selection models comes at a cost in the form of an increased number of parameters. 
While codon substitution models with uniform stationary states typically have $<$20 parameters, mutation-selection models must specify 19 parameters for \emph{each} site (the stationary state is for 20 amino acids whose frequencies are constrained to sum to one).
This corresponds to $19\times L$ parameters for a protein of length $L$, or $\sim 10^4$ parameters for a typical size protein.
As with all parameter-rich models, it is important to obtain values for these parameters using some method that avoids overfitting \skhcomment{cite rodrigue paper}.
Here we will primarily use experimentally informed codon models (ExpCM)~\citep{bloom2014experimentally, hilton2017phydms, bloom2017identification} which define values for these parameters \textit{a priori} from deep mutational scanning experiments~\citep{araya2011,fowler2010high} so they do not need to be fit from phylogenetic data.
Therefore, the number of remaining free parameters that are fit from the phylogenetic data for an ExpCM are similar to a non-site-specific substitution model.
Alternative strategies of obtaining parameters for site-specific stationary states via Bayesian\skhcomment{cite} or maximum-likelihood estimation\skhcomment{cite} are discussed in the last section of the Results.

Finally, these two strategies to account for purifying selection, $\Gamma\omega$ rate variation and site-specific stationary states, are not mutually exclusive. 
Some proteins may be better modeled as a combination of the two. 
Therefore, we will use models from the GY94 (uniform stationary state) and ExpCM (site-specific stationary state) families with and without $\Gamma\omega$ rate variation (\ref{fig:model_feature}C) to examine the effects of each strategy on branch length estimation.

>>>>

\begin{abstract}
\noindent  
Molecular phylogenetics is used to estimate the time since the divergence of modern gene sequences.
Such phylogenetic techniques often estimate substantially shallower divergence times than other methods.
For instance, in the case of viruses there is independent evidence that molecular phylogenetics can underestimate deep divergence times.  
This discrepancy is thought to be caused in part by inadequate models of purifying selection leading to branch-length underestimation.
Here we show that substitution models informed by experimental measurements of the purifying selection \skhcomment{due to site-specific amino-acid preferences} lengthen the deep branches on phylogenies of influenza virus hemagglutinin.
This deepening of branch lengths is due to better modeling of the stationary state of the substitution models, and is independent of the branch-length-extension that results from modeling site-to-site variation in substitution rate.
The deepening of branch lengths from experimentally informed site-specific substitution models is similar to that achieved by other approaches that allow the stationary state to vary across sites \skhcomment{repeat sentence start}.
However, the improvements from these site-specific models are limited by the inherent tension between the enhanced accuracy of accounting for site-specific amino-acid preferences and the fact that these preferences shift over long evolutionary times.
Overall, our work underscores the importance of modeling how site-specific purifying selection affects the stationary state when estimating deep divergence times. 
\end{abstract}

@@

Proteins evolve under purifying selection to maintain their structure and function. 
This purifying selection is not homogenous across sites in a protein.
It is also not homogenous among the different amino acids at a given site.
For instance, some protein sites strongly prefer hydrophobic amino acids, others may be constrained to just one or a few amino acids, and yet others may tolerate many amino acids.
In general, these constraints are highly idiosyncratic among sites, and so pose a challenge for phylogenetic substitution models.

The most common strategy to model rate variation is to allow the rate of non-synonymous change to vary among sites following some statistical distribution.
For example, in GY94+$\Gamma\omega$ (M5 of \cite{yang2000codon}), the implied dN/dS parameter, $\omega$, follows a gamma distribution (\ref{fig:model_feature}A).
Under GY94+$\Gamma\omega$ and other such models, some sites have a higher rate of non-synonymous change than others. 
However, this rate is agnostic to the amino-acid identities themselves, and so all non-synonymous changes at at given site are treated equally.

In contrast, so-called ``mutation-selection'' models~\citep{halpern1998evolutionary} account for purifying selection by explicitly defining a different set of amino-acid preferences at each site in the protein. 
This more mechanistic formulation, without an arbitrary statistical distribution, results in a site-specific stationary state (\ref{fig:model_feature}B). 
These models better capture the site-to-site differences in amino-acid composition that is an obvious features of real proteins, and indeed they generally better describe actual evolution than models with only rate variation~\citep{lartillot2004bayesian, le2008phylogenetic, rodrigue2010mutation,hilton2017phydms,bloom2014experimentally}.
However, the specificity of mutation-selection models comes at a cost in the form of an increased number of parameters. 
While codon substitution models with uniform stationary states typically have $<$20 parameters, mutation-selection models must specify 19 parameters for \emph{each} site (the stationary state is for 20 amino acids whose frequencies are constrained to sum to one).
This corresponds to $19\times L$ parameters for a protein of length $L$, or $\sim 10^4$ parameters for a typical size protein.
As with all parameter-rich models, it is important to obtain values for these parameters using some method that avoids overfitting \skhcomment{cite rodrigue paper}.
Here we will primarily use experimentally informed codon models (ExpCM)~\citep{bloom2014experimentally, hilton2017phydms, bloom2017identification} which define values for these parameters \textit{a priori} from deep mutational scanning experiments~\citep{araya2011,fowler2010high} so they do not need to be fit from phylogenetic data.
Therefore, the number of remaining free parameters for an ExpCM are similar to a non-site-specific substitution model.
Alternative strategies of obtaining parameters for site-specific stationary states via Bayesian\skhcomment{cite} or maximum-likelihood estimation\skhcomment{cite} are discussed in the last section of the Results.

Finally, these two strategies to account for purifying selection, $\Gamma\omega$ rate variation and site-specific stationary states, are not mutually exclusive. 
Some proteins may be better modeled as a combination of the two. 
Therefore, we will use models from the GY94 (uniform stationary state) and ExpCM (site-specific stationary state) families with and without $\Gamma\omega$ rate variation (\ref{fig:model_feature}C) to dissect the effect of each model of purifying selection on branch length estimation.
